Trends from 30-Year Observations of Downward Solar Irradiance in Thessaloniki, Greece

### 2024-01-01 MDPI_submission.Rmd
Commit 2024-01-01 18:12

## order columns
LT1 <- LT[,c(
      "Sky conditions",
      "Trend [%/year]",
      "Trend S.E.",
      "Pearson cor.",
      "Trend [W/year]",
      "Days"
      )]
# LT1 <- LT

# signif(LT1$`Trend [%/year]` , 3)
# round(LT1$`Trend [%/year]` , 3)
# format(LT1$`Trend [%/year]` , 3, drop0trailing = T)
# as.character(format(LT1$`Trend [%/year]` , 3))

## order columns
# LT2 <- LT[,c(
#       "Sky conditions",
#       "Trend [%/year]",
#       "Statistical signif. [%]",
#       "Days with data"
#       )]
LT2 <- LT
```

We calculated the linear trends of SDR from the departures of the mean daily values from the daily climatology and for the three sky conditions.
These are presented in Table\ \ref{tab:trendtable}, which also contains the $2\sigma$ standard error, the Pearson's correlation coefficient R, and the trend in absolute units.
In Figure\ \ref{fig:trendALL}, we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in Appendix (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
The linear trends in all three datasets are positive and around
$0.4\,\%/y$ for all-sky and cloudy-sky conditions, whereas for clear-skies the
trend is much smaller (\~$`r signif(cleartrend,1)`\,\%/y$). 
The linear trends were calculated
taking into account the autocorrelation of the time series, and all three
are statistically significant at least at the $95\,\%$ confidence level, as
they are larger than the corresponding $2\sigma$ errors, despite the small values of R, which is due to the large variability of the daily values.
The clear-sky trend is
very small, suggesting a small effect from aerosols and water vapor, which
are the dominant factors of the SDR variability [@Fountoulakis2016; @Siomos2018; @Yu2022].
In contrast, the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or cloud optical depth. 
Lack of continuous observations of cloud optical thickness that could support these findings does not allow drawing
firm conclusions. However, there are indications that the total
cloud cover as inferred from the ERA5 analysis for the grid point of
Thessaloniki is decreasing over the period of study.
From the difference between all-sky and clear-sky SDR trends, expressed in $W/m^2/y$ using the long-term mean of the respective datasets, the radiative effect of clouds is estimated to $0.96\,W/m^2/y$. This estimate is similar to the cloud radiative forcing of $1.22\,W/m^2/y$ reported for Granada, Spain [@Lozano2023].

@@

```{r trendtable, echo=F}
# names(LT1)
# "Sky conditions", "Trend [%/year]", "Trend S.E.", "Pearson cor.", "Trend [W/year]", "Days" 
x_latex <- knitr::kable(LT1,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky conditions", "Trend [\\%/year]", "Trend S.E. ($2\\sigma$)", "Pearson correl.", "Trend [W/m\\textsuperscript{2}/year]", "Days"),
                        format.args = list(digits = 3),
                        linesep     = "",
                        align       = "crrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        position    = "H",
                        drop0trailing = T,
                        caption     = paste0("Trends in SDR daily means for different sky conditions for the period ", year(min(ALL_1_daily_DESEAS$Date)), "--", year(max(ALL_1_daily_DESEAS$Date)), "."))

# x_latex %>%
#     kable_styling(
#         latex_options = c("hold_position")
#     )
x_latex %>%
    kable_styling(full_width = TRUE) %>%
    column_spec(1, width = "8em")
```

```{r trendALL, echo=F, out.width='.75\\linewidth', fig.cap=paste0("Anomalies (\\%) of the daily all-sky SDR from the climatological mean for the period ", year(min(ALL_1_daily_DESEAS$Date)), " -- ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line is the long term linear trend.")}
knitr::include_graphics("./images/LongtermTrends-1.png")
```

Although the year-to-year variability of the anomalies (Figure \ref{fig:trendALL} and Figures\ \ref{fig:trendCLEAR} and \ref{fig:trendCLOUD} in Appendix) shows a rather homogeneous behavior, plots of the cumulative sums (CUSUM) [@Regier2019] of the anomalies can reveal different structures in the records of all three sky conditions.
For time series with a uniform trend, we would expect
the CUSUMs of the anomalies to have a symmetric 'V' shape centered
around the middle of the series. This would indicate that the anomalies
are evenly distributed around the climatological mean, and, for a
positive uniform trend, the first half is below and the second half
above the climatological mean.
In our case, there is a more complex
evolution of the anomalies.
For all skies (Figure\ \ref{fig:cusummonth-1}),
we observe three
rather distinct periods: (a) a downward part between the start of the
datasets and about 2000, denoting that all anomalies are negative, thus
below the climatology; (b) a relatively steady part lasting for almost 20
years suggesting little variability in SDR anomalies; and (c) a steep
upward part to the present, indicating anomalies above the climatology. The
CUSUMs for cloudy skies (Figure\ \ref{fig:cusummonth-3}) show a similar behavior with some short-term
differences that do not change the overall pattern.
For clear skies
(Figure\ \ref{fig:cusummonth-2}), a monotonic downward tendency is evident until 2004,
suggesting that the anomalies are all negative. After 2004, the anomalies
turn positive at a fast rate for about five years and at a slower
rate thereafter.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
        \subfloat[All skies.\label{fig:cusummonth-1}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
        \subfloat[Clear skies.\label{fig:cusummonth-2}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
        \subfloat[Cloudy skies.\label{fig:cusummonth-3}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}\hfill
        }
\caption{Cumulative sum plots of the monthly SDR anomalies in (\%) for different sky conditions.}\label{fig:cusummonth}
\end{adjustwidth}
\end{figure}

<!-- ```{r cusummonth, echo = FALSE, fig.cap='Cumulative sum plots of the monthly SDR anomalies in (%) for different sky conditions.', fig.subcap=c('All skies.', 'Clear skies.', 'Cloudy skies.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics(c( -->
<!--     "./images/CumulativeMonthlyCuSum-1.png",    # ALL sky -->
<!--     "./images/CumulativeMonthlyCuSum-5.png",    # CLEAR sky -->
<!--     "./images/CumulativeMonthlyCuSum-9.png"))   # Cloud sky -->
<!-- ``` -->

@@

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}

<!-- ```{r cusumnotrendmonthly, echo = FALSE, fig.cap='Cumulative sum plots of monthly SDR anomalies in (%) for different sky conditions after removing the long-term linear trend.', fig.subcap=c('All skies.', 'Clear skies.', 'Cloudy skies.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics(c( -->
<!--     "./images/CumulativeMonthlyCuSumNOtrend-1.png",    # ALL sky -->
<!--     "./images/CumulativeMonthlyCuSumNOtrend-5.png",    # CLEAR sky -->
<!--     "./images/CumulativeMonthlyCuSumNOtrend-9.png"))   # Cloud sky -->
<!-- ``` -->

<!-- ```{r cusum, echo = FALSE, fig.cap='Running cumulative sum plots of the daily SDR.', fig.subcap=c('All-sky conditions.', 'Clear-sky conditions.', 'Cloud-sky conditions.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics(c( -->
<!--     "./images/CumulativeDailyCuSum-1.png",    # ALL sky -->
<!--     "./images/CumulativeDailyCuSum-5.png",    # CLEAR sky -->
<!--     "./images/CumulativeDailyCuSum-9.png"))   # Cloud sky -->
<!-- ``` -->

<!-- ```{r cusumnotrend, echo = FALSE, fig.cap='Running cumulative sum plots of the daily SDR after removing the long term trend from the daily values.', fig.subcap=c('All-skies.', 'Clear-skies.', 'Cloud-skies.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics(c( -->
<!--     "./images/CumulativeDailyCuSumNOtrend-1.png",    # ALL sky -->
<!--     "./images/CumulativeDailyCuSumNOtrend-5.png",    # CLEAR sky -->
<!--     "./images/CumulativeDailyCuSumNOtrend-9.png"))   # Cloud sky -->
<!-- ``` -->

<!-- ```{r cusumSTD, echo = FALSE, fig.cap='two plots', fig.subcap=c('one plot', 'the other one'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics(c( -->
<!--     "./images/CumulativeDailyCuSumSTD-1.png",    # ALL sky -->
<!--     "./images/CumulativeDailyCuSumSTD-2.png",    # CLEAR sky -->
<!--     "./images/CumulativeDailyCuSumSTD-3.png"))   # Cloud sky -->
<!-- ``` -->

>>>>

## order columns
LT1 <- LT[,c(
      "Sky conditions",
      "Trend [%/year]",
      "Trend S.E.",
      "Pearson cor.",
      "Trend [W/year]",
      "Days"
      )]
```

We calculated the linear trends of SDR from the departures of the mean daily values from the daily climatology and for the three sky conditions.
These are presented in Table\ \ref{tab:trendtable}, which also contains the $2\sigma$ standard error, the Pearson's correlation coefficient R, and the trend in absolute units.
In Figure\ \ref{fig:trendALL}, we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in Appendix \ref{appendix} (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
The linear trends in all three datasets are positive and around
$0.4\,\%/y$ for all-sky and cloudy-sky conditions, whereas for clear-skies the
trend is much smaller (\~$`r signif(cleartrend,1)`\,\%/y$). 
The linear trends were calculated
taking into account the autocorrelation of the time series, and all three
are statistically significant at least at the $95\,\%$ confidence level, as
they are larger than the corresponding $2\sigma$ errors, despite the small values of R, which is due to the large variability of the daily values.
The clear-sky trend is
very small, suggesting a small effect from aerosols and water vapor, which
are the dominant factors of the SDR variability [@Fountoulakis2016; @Siomos2018; @Yu2022].
In contrast, the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or cloud optical depth. 
Lack of continuous observations of cloud optical thickness that could support these findings does not allow drawing
firm conclusions. However, there are indications that the total
cloud cover as inferred from the ERA5 analysis for the grid point of
Thessaloniki is decreasing over the period of study.
From the difference between all-sky and clear-sky SDR trends, expressed in $W/m^2/y$ using the long-term mean of the respective datasets, the radiative effect of clouds is estimated to $0.96\,W/m^2/y$. This estimate is similar to the cloud radiative forcing of $1.22\,W/m^2/y$ reported for Granada, Spain [@Lozano2023].

@@

```{r trendtable, echo=F}
# names(LT1)
# "Sky conditions", "Trend [%/year]", "Trend S.E.", "Pearson cor.", "Trend [W/year]", "Days" 
x_latex <- knitr::kable(LT1,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky Conditions", "Trend [\\%/year]", "Trend S.E. ($2\\sigma$)", "Pearson Correl.", "Trend [W/m\\textsuperscript{2}/year]", "Days"),
                        format.args = list(digits = 3),
                        linesep     = "",
                        align       = "crrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        position    = "H",
                        drop0trailing = T,
                        caption     = paste0("Trends in SDR daily means for different sky conditions for the period ", year(min(ALL_1_daily_DESEAS$Date)), "--", year(max(ALL_1_daily_DESEAS$Date)), "."))

x_latex                              %>%
    row_spec(0,bold=TRUE)            %>%
    kable_styling(full_width = TRUE) %>%
    column_spec(1, width = "8em")
```

```{r trendALL, echo=F, out.width='.75\\linewidth', fig.cap=paste0("Anomalies (\\%) of the daily all-sky SDR from the climatological mean for the period ", year(min(ALL_1_daily_DESEAS$Date)), "--", year(max(ALL_1_daily_DESEAS$Date)), ". The black line is the long-term linear trend.")}
knitr::include_graphics("./images/LongtermTrends-1.png")
```

Although the year-to-year variability of the anomalies (Figures\ \ref{fig:trendALL}, \ref{fig:trendCLEAR} and \ref{fig:trendCLOUD} in Appendix \ref{appendix}) shows a rather homogeneous behavior, plots of the cumulative sums (CUSUM) [@Regier2019] of the anomalies can reveal different structures in the records of all three sky conditions.
For time series with a uniform trend, we would expect
the CUSUMs of the anomalies to have a symmetric 'V' shape centered
around the middle of the series. This would indicate that the anomalies
are evenly distributed around the climatological mean, and, for a
positive uniform trend, the first half is below and the second half
above the climatological mean.
In our case, there is a more complex
evolution of the anomalies.
For all skies (Figure\ \ref{fig:cusummonth}a),
we observe three
rather distinct periods: (a) a downward part between the start of the
datasets and about 2000, denoting that all anomalies are negative, thus
below the climatology; (b) a relatively steady part lasting for almost 20
years, suggesting little variability in SDR anomalies; and (c) a steep
upward part to the present, indicating anomalies above the climatology. The
CUSUMs for cloudy skies (Figure\ \ref{fig:cusummonth}c) show a similar behavior with some short-term
differences that do not change the overall pattern.
For clear skies
(Figure\ \ref{fig:cusummonth}b), a monotonic downward tendency is evident until 2004,
suggesting that the anomalies are all negative. After 2004, the anomalies
turn positive at a fast rate for about five years and at a slower
rate thereafter.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}
        }
    \end{adjustwidth}
\caption{Cumulative sum plots of the monthly SDR anomalies in (\%) for different sky conditions: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusummonth}
\end{figure}

\vspace{-6pt}

@@

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}

### 2024-01-01 MDPI_submission.tex
Commit 2024-01-01 18:12

We calculated the linear trends of SDR from the departures of the mean
daily values from the daily climatology and for the three sky
conditions. These are presented in Table~\ref{tab:trendtable}, which
also contains the \(2\sigma\) standard error, the Pearson's correlation
coefficient R, and the trend in absolute units. In
Figure~\ref{fig:trendALL}, we present only the time series under all-sky
conditions; the plots for clear-sky and cloud-sky conditions, are shown
in Appendix (Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). In
the studied period, there is no significant break or change in the
variability pattern of the time series. The linear trends in all three
datasets are positive and around \(0.4\,\%/y\) for all-sky and
cloudy-sky conditions, whereas for clear-skies the trend is much smaller
(\textasciitilde{}\(0.1\,\%/y\)). The linear trends were calculated
taking into account the autocorrelation of the time series, and all
three are statistically significant at least at the \(95\,\%\)
confidence level, as they are larger than the corresponding \(2\sigma\)
errors, despite the small values of R, which is due to the large
variability of the daily values. The clear-sky trend is very small,
suggesting a small effect from aerosols and water vapor, which are the
dominant factors of the SDR variability
\citep{Fountoulakis2016, Siomos2018, Yu2022}. In contrast, the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or cloud optical depth. Lack of continuous
observations of cloud optical thickness that could support these
findings does not allow drawing firm conclusions. However, there are
indications that the total cloud cover as inferred from the ERA5
analysis for the grid point of Thessaloniki is decreasing over the
period of study. From the difference between all-sky and clear-sky SDR
trends, expressed in \(W/m^2/y\) using the long-term mean of the
respective datasets, the radiative effect of clouds is estimated to
\(0.96\,W/m^2/y\). This estimate is similar to the cloud radiative
forcing of \(1.22\,W/m^2/y\) reported for Granada, Spain
\citep{Lozano2023}.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993--2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
Sky conditions & Trend [\%/year] & Trend S.E. ($2\sigma$) & Pearson correl. & Trend [W/m\textsuperscript{2}/year] & Days\\
\midrule
All skies & 0.380 & 0.120 & 0.091 & 1.460 & 10251\\
Clear skies & 0.097 & 0.033 & 0.140 & 0.501 & 2684\\
Cloudy skies & 0.410 & 0.180 & 0.081 & 1.180 & 4937\\
\bottomrule
\end{tabu}
\end{table}

@@

\caption{Anomalies (\%) of the daily all-sky SDR from the climatological mean for the period 1993 -- 2023. The black line is the long term linear trend.}\label{fig:trendALL}
\end{figure}

Although the year-to-year variability of the anomalies (Figure
\ref{fig:trendALL} and Figures~\ref{fig:trendCLEAR} and
\ref{fig:trendCLOUD} in Appendix) shows a rather homogeneous behavior,
plots of the cumulative sums (CUSUM) \citep{Regier2019} of the anomalies
can reveal different structures in the records of all three sky
conditions. For time series with a uniform trend, we would expect the
CUSUMs of the anomalies to have a symmetric `V' shape centered around
the middle of the series. This would indicate that the anomalies are
evenly distributed around the climatological mean, and, for a positive
uniform trend, the first half is below and the second half above the
climatological mean. In our case, there is a more complex evolution of
the anomalies. For all skies (Figure~\ref{fig:cusummonth-1}), we observe
three rather distinct periods: (a) a downward part between the start of
the datasets and about 2000, denoting that all anomalies are negative,
thus below the climatology; (b) a relatively steady part lasting for
almost 20 years suggesting little variability in SDR anomalies; and (c)
a steep upward part to the present, indicating anomalies above the
climatology. The CUSUMs for cloudy skies (Figure~\ref{fig:cusummonth-3})
show a similar behavior with some short-term differences that do not
change the overall pattern. For clear skies
(Figure~\ref{fig:cusummonth-2}), a monotonic downward tendency is
evident until 2004, suggesting that the anomalies are all negative.
After 2004, the anomalies turn positive at a fast rate for about five
years and at a slower rate thereafter.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
        \subfloat[All skies.\label{fig:cusummonth-1}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
        \subfloat[Clear skies.\label{fig:cusummonth-2}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
        \subfloat[Cloudy skies.\label{fig:cusummonth-3}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}\hfill
        }
\caption{Cumulative sum plots of the monthly SDR anomalies in (\%) for different sky conditions.}\label{fig:cusummonth}
\end{adjustwidth}
\end{figure}

@@

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}

>>>>

We calculated the linear trends of SDR from the departures of the mean
daily values from the daily climatology and for the three sky
conditions. These are presented in Table~\ref{tab:trendtable}, which
also contains the \(2\sigma\) standard error, the Pearson's correlation
coefficient R, and the trend in absolute units. In
Figure~\ref{fig:trendALL}, we present only the time series under all-sky
conditions; the plots for clear-sky and cloud-sky conditions, are shown
in Appendix \ref{appendix} (Figures~\ref{fig:trendCLEAR} and~
\ref{fig:trendCLOUD}). In the studied period, there is no significant
break or change in the variability pattern of the time series. The
linear trends in all three datasets are positive and around
\(0.4\,\%/y\) for all-sky and cloudy-sky conditions, whereas for
clear-skies the trend is much smaller (\textasciitilde{}\(0.1\,\%/y\)).
The linear trends were calculated taking into account the
autocorrelation of the time series, and all three are statistically
significant at least at the \(95\,\%\) confidence level, as they are
larger than the corresponding \(2\sigma\) errors, despite the small
values of R, which is due to the large variability of the daily values.
The clear-sky trend is very small, suggesting a small effect from
aerosols and water vapor, which are the dominant factors of the SDR
variability \citep{Fountoulakis2016, Siomos2018, Yu2022}. In contrast,
the large positive trend of SDR under cloudy skies can be attributed to
reduction in cloud cover and/or cloud optical depth. Lack of continuous
observations of cloud optical thickness that could support these
findings does not allow drawing firm conclusions. However, there are
indications that the total cloud cover as inferred from the ERA5
analysis for the grid point of Thessaloniki is decreasing over the
period of study. From the difference between all-sky and clear-sky SDR
trends, expressed in \(W/m^2/y\) using the long-term mean of the
respective datasets, the radiative effect of clouds is estimated to
\(0.96\,W/m^2/y\). This estimate is similar to the cloud radiative
forcing of \(1.22\,W/m^2/y\) reported for Granada, Spain
\citep{Lozano2023}.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993--2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
\textbf{Sky Conditions} & \textbf{Trend [\%/year]} & \textbf{Trend S.E. ($2\sigma$)} & \textbf{Pearson Correl.} & \textbf{Trend [W/m\textsuperscript{2}/year]} & \textbf{Days}\\
\midrule
All skies & 0.380 & 0.120 & 0.091 & 1.460 & 10251\\
Clear skies & 0.097 & 0.033 & 0.140 & 0.501 & 2684\\
Cloudy skies & 0.410 & 0.180 & 0.081 & 1.180 & 4937\\
\bottomrule
\end{tabu}
\end{table}

@@

\caption{Anomalies (\%) of the daily all-sky SDR from the climatological mean for the period 1993--2023. The black line is the long-term linear trend.}\label{fig:trendALL}
\end{figure}

Although the year-to-year variability of the anomalies
(Figures~\ref{fig:trendALL}, \ref{fig:trendCLEAR} and
\ref{fig:trendCLOUD} in Appendix \ref{appendix}) shows a rather
homogeneous behavior, plots of the cumulative sums (CUSUM)
\citep{Regier2019} of the anomalies can reveal different structures in
the records of all three sky conditions. For time series with a uniform
trend, we would expect the CUSUMs of the anomalies to have a symmetric
`V' shape centered around the middle of the series. This would indicate
that the anomalies are evenly distributed around the climatological
mean, and, for a positive uniform trend, the first half is below and the
second half above the climatological mean. In our case, there is a more
complex evolution of the anomalies. For all skies
(Figure~\ref{fig:cusummonth}a), we observe three rather distinct
periods: (a) a downward part between the start of the datasets and about
2000, denoting that all anomalies are negative, thus below the
climatology; (b) a relatively steady part lasting for almost 20 years,
suggesting little variability in SDR anomalies; and (c) a steep upward
part to the present, indicating anomalies above the climatology. The
CUSUMs for cloudy skies (Figure~\ref{fig:cusummonth}c) show a similar
behavior with some short-term differences that do not change the overall
pattern. For clear skies (Figure~\ref{fig:cusummonth}b), a monotonic
downward tendency is evident until 2004, suggesting that the anomalies
are all negative. After 2004, the anomalies turn positive at a fast rate
for about five years and at a slower rate thereafter.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}
        }
    \end{adjustwidth}
\caption{Cumulative sum plots of the monthly SDR anomalies in (\%) for different sky conditions: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusummonth}
\end{figure}

\vspace{-6pt}

@@

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}
